An Adaptive Algorithm Employing Continuous Linear Functionals

### 2019-05-28 AutoAlgorithmLinearFunctionalDataVerF.tex
fred edits

\begin{algo}\label{algo2}
Given $a$, $b$, the sequence $\bsn$, the cone $\calC$, the input function $f \in \calC$, and the absolute error tolerance $\varepsilon$, set $j=1$.
\begin{description}
\item[Step 1.] Compute $\sigma_{j}(f)$ as defined in \eqref{sumdef}.
\item[Step 2.] Check whether $j$ is large enough to satisfy the error tolerance, i.e.,
    \begin{equation*}\label{covcrit}
          \sigma_{j}(f) \le \frac{\varepsilon\sqrt{1 - b^2}}{ab} .
    \end{equation*}
    If this is true, then return $\widetilde{A}(f,\varepsilon) = A_{n_{j}}(f)$, where $A_n$ is defined in \eqref{optAdef}, and terminate the algorithm.
\item[Step 3.] Otherwise, increase $j$ by $1$ and return to Step $1$.
\end{description}
\end{algo}
\yuhannote{Please check the result of Theorem 1} \frednote{Yuhan, please put $\rho$ and $\varepsilon$ on left.}\yuhannote{Done. Please double check}
\begin{theorem}\label{thm:compcost}
The algorithm, $\widetilde{A}$, defined in Algorithm \ref{algo2} lies in $\calA(\calC)$ and has computational cost $\textup{cost}(\widetilde{A},f,\varepsilon)=n_{j^*}$, where $j^*$ is defined implicitly by the inequalities 
\begin{equation} \label{eq:OurAlgjstar}
j^* = \min\left \{ j \in \bbN : \sigma_{j}(f) \le \frac{\varepsilon\sqrt{1 - b^2}}{ab}  \right\}.
\end{equation}
Moreover, $\textup{cost}(\widetilde{A},\calC,\varepsilon,\rho) \frednote{=} n_{j^\dagger}$, where $j^\dagger$ satisfies the following upper bound:
\begin{equation} \label{jdagger}
j^\dagger \le \min \left \{j \in \bbN : F(j) \ge \frac{\rho^2 a^4}{\varepsilon^2(1 - b^2)} \right\},
\end{equation}
% \begin{equation} \label{jdagger}
% j^\dagger \le \min \left \{j \in \bbN : \frac{\rho^2}{\varepsilon^2} \le \frac{(1 - b^2)}{a^4} F(j) \right\},
% \end{equation}
where $F$ is the strictly increasing function defined as
\begin{equation} \label{jsum}
F(j): = \sum_{k=0}^{j-1} \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}, \qquad j \in \N.
\end{equation}
\end{theorem}

@@

Although two test functions $f_\pm$ yield the same approximate solution, they have different true solutions.  In particular,
\begin{align*}
\varepsilon &\ge \max \bigl\{\DHJRnorm[\calG]{S(f_+) - A^*(f_+,\varepsilon)}, \DHJRnorm[\calG]{S(f_-) - A^*(f_-,\varepsilon)} \bigr\} \\
&\ge \frac 12 \bigl[\DHJRnorm[\calG]{S(f_+) - A^*(f,\varepsilon)} + \DHJRnorm[\calG]{S(f_-) - A^*(f,\varepsilon)}  \bigr] \\
&\qquad \qquad \text{since } A^*(f_\pm,\varepsilon) = A^*(f,\varepsilon) \\
&\ge \frac 12 \DHJRnorm[\calG]{S(f_+) - S(f_-)} \quad \text{by the triangle inequality}\\
&\ge \frac 12 \DHJRnorm[\calG]{S(f_+ - f_-)} \quad \text{since $S$ is linear}\\
&= \eta \DHJRnorm[\calG]{S(u)}.
\end{align*}
Thus, we have \yuhannote{Please check the proof below}
\begin{align*}
\varepsilon^2  & \ge \eta^2 \DHJRnorm[\calG]{S(u)}^2= 
\eta^2 \sum_{k=0}^{j} \DHJRbignorm[\calG]{S(u^{(k)})}^2  \frac{b^{2(k-j)}}{\lambda_{n_k}^2}\\
& \ge \eta^2 
\sum_{k=0}^{j} \DHJRbignorm[\calF]{u^{(k)}}^2 
b^{2(k-j)} 
\qquad {\color{red} \text{since } \DHJRbignorm[\calG]{S(u^{(k)})} \ge \lambda_{n_k} \DHJRbignorm[\calF]{u^{(k)}} }\\
& \ge  \eta^2 b^{2(k^{*}-j)} \qquad \text{ where } k^* = \argmax_{0 \le k \le j} \DHJRbignorm[\calF]{u^{(k)}} \\
& \ge \eta^2 = \frac{(a-1)^2c^2}{(a+1)^2R^2} 
\qquad {\color{red} \text{since } b < 1}\\
&
=\frac{(a-1)^2 \rho^2}{(a+1)^2R^2}\left[\left(1+\frac{(a-1)^2}{(a+1)^{2} R^2 }\right)\sum_{k=0}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}\right]^{-1} 
\qquad {\color{red} \text{by \eqref{DHJRetaDef}}}\\
& =  \rho^2 \left[\left(\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right)\sum_{k=0}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}\right]^{-1} \\
& = \rho^2 \left[\left(\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right) b^2F(j+1)\right]^{-1} 
\qquad {\color{red} \text{by \eqref{jsum}}}.
\end{align*}
\frednote{This inequality is equivalent to 
\begin{equation}
    F(j+1) \ge  \frac{\rho^2}{\varepsilon^2 b^2}
\left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right]^{-1}
\end{equation}
}

This lower bound must be satisfied by $j$ to be consistent with the assumption $\textup{comp}(\calA(\calC),\varepsilon,\rho) \le n_j - 2$.  Thus, for any $j$ violating this inequality it follows that $\textup{comp}(\calA(\calC),\varepsilon,\rho) \ge n_j - 1$.  This implication provides a lower bound on $\textup{comp}(\calA(\calC),\varepsilon,\rho)$.
\end{proof}

The next step is to show that the cost of our algorithm is essentially no worse than that of the optimal algorithm.
\yuhannote{The proof of theorem 3 needs to be checked.}
\begin{theorem}
\label{thm:CostNoWorse}
Under assumption \eqref{lambdaRatio} $\textup{cost}(\widetilde{A},\calC, \varepsilon,\rho)$ is essentially no worse than $\textup{comp}(\calA(\calC),\varepsilon,\rho)$. 
\end{theorem}
\begin{proof} Let
\begin{equation} \label{newomegadef}
    \omega = \sqrt{\frac{(1 - b^2)b^4}{a^4(1 + b^2R^2 + b^4R^4 + b^6R^6)}\left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right]^{-1}},
\end{equation}
and note that it does not depend on $\rho$ or $\varepsilon$ but only on the definition of $\calC$. 
For any positive $\rho$ and $\varepsilon$, Theorem \ref{thm:compcost} says that $\textup{cost}(\widetilde{A},\calC,\varepsilon,\rho) \le n_{j^\dagger(\varepsilon,\rho)}$, where 
\[
j^\dagger(\varepsilon,\rho) \le \min \left \{j \in \bbN : F(j) \ge \frac{\rho^2 a^4}{\varepsilon^2 (1-b^2)}  \right\}. \]
According to \eqref{lambdaRatio}, for any $\ell \in \mathbb{N}$
\begin{align*}
 (b^2R^2)^\ell F(j)
& =(b^2R^2)^\ell \sum_{k=0}^{j-1} \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2} 
 = (R^2)^\ell\sum_{k=\ell}^{j+\ell-1} \frac{b^{2(k-j)}}{\lambda_{n_{k-\ell}}^2} \\
& \geq (R^2)^{\ell-1}\sum_{k=\ell}^{j+\ell-1} \frac{b^{2(k-j)}}{\lambda_{n_{k-\ell+1}}^2} \\
 & \geq \sum_{k=\ell}^{j+\ell-1} \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2} \\
& \geq \frac{b^{2(\ell-1)}}{\lambda_{n_{j+\ell-1}}^2},
\end{align*}
thus 
\begin{align*} 
  \frac{(1 - b^2)}{a^4} F(j) 
&\ge  \frac{(1 - b^2)}{a^4(1 + b^2R^2 + b^4R^4 + b^6R^6)} \sum_{k=0}^{j+2} \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}  \\
& = \frac{(1 - b^2)}{a^4(1 + b^2R^2 + b^4R^4 + b^6R^6)} b^6 F(j+3)\\
&=  \omega^2\left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right] b^2 F(j+3) \qquad \text{by \eqref{newomegadef}.}
\end{align*}
Hence,
\begin{align*} 
j^\dagger(\varepsilon,\rho) &\le \min \left \{j \in \bbN : F(j) \ge \frac{\rho^2 a^4}{\varepsilon^2 (1-b^2)}   \right\} \\
& =
 \min \left \{j \in \bbN : \frac{(1 - b^2)}{a^4} F(j) \ge \frac{\rho^2 }{\varepsilon^2 }   \right\} \\
& \le \min \left \{j \in \bbN : \omega^2\left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right] b^2 F(j+3) \ge \frac{\rho^2}{\varepsilon^2}  \right\} \\
& = \min \left \{j \in \bbN :  F(j+3) \ge \frac{\rho^2}{\omega^2 \varepsilon^2 b^2} \left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right]^{-1} \right\} \\
& = j^{\ddagger}(\omega\varepsilon,\rho) - 1.
\end{align*}
By the argument above, it follows that $j^\ddagger -1 \ge j^\dagger$ and by Theorem 2 so
\[
\textup{comp}(\calA(\calC),\omega\varepsilon,\rho) \ge n_{j^{\ddagger}} -1 \ge n_{j^{\ddagger}-1} \ge  n_{j^\dagger}\ge \textup{cost}(\widetilde{A},\calC,\varepsilon,\rho) .\]
Therefore, our algorithm is essentially no more costly than the optimal algorithm. 
\end{proof}

>>>>

\begin{algo}\label{algo2}
Given $a$, $b$, the sequence $\bsn$, the cone $\calC$, the input function $f \in \calC$, and the absolute error tolerance $\varepsilon$, set $j=1$.
\begin{description}
\item[Step 1.] Compute $\sigma_{j}(f)$ as defined in \eqref{sumdef}.
\item[Step 2.] Check whether $j$ is large enough to satisfy the error tolerance, i.e.,
    \begin{equation*}\label{covcrit}
          \sigma_{j}(f) \le \frac{\varepsilon\sqrt{1 - b^2}}{ab} .
    \end{equation*}
    If this is true, then return $\widetilde{A}(f,\varepsilon) = A_{n_{j}}(f)$, where $A_n$ is defined in \eqref{optAdef}, and terminate the algorithm.
\item[Step 3.] Otherwise, increase $j$ by $1$ and return to Step $1$.
\end{description}
\end{algo}
\yuhannote{Please check the result of Theorem 1} 
\begin{theorem}\label{thm:compcost}
The algorithm, $\widetilde{A}$, defined in Algorithm \ref{algo2} lies in $\calA(\calC)$ and has computational cost $\textup{cost}(\widetilde{A},f,\varepsilon)=n_{j^*}$, where $j^*$ is defined implicitly by the inequalities 
\begin{equation} \label{eq:OurAlgjstar}
j^* = \min\left \{ j \in \bbN : \sigma_{j}(f) \le \frac{\varepsilon\sqrt{1 - b^2}}{ab}  \right\}.
\end{equation}
Moreover, $\textup{cost}(\widetilde{A},\calC,\varepsilon,\rho) \frednote{=} n_{j^\dagger}$, where $j^\dagger$ satisfies the following upper bound:
\begin{equation} \label{jdagger}
j^\dagger \le \min \left \{j \in \bbN : F(j) \ge \frac{\rho^2 a^4}{\varepsilon^2(1 - b^2)} \right\},
\end{equation}
% \begin{equation} \label{jdagger}
% j^\dagger \le \min \left \{j \in \bbN : \frac{\rho^2}{\varepsilon^2} \le \frac{(1 - b^2)}{a^4} F(j) \right\},
% \end{equation}
where $F$ is the strictly increasing function defined as
\begin{equation} \label{jsum}
F(j): = \sum_{k=0}^{j-1} \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}, \qquad j \in \N.
\end{equation}
\end{theorem}

@@

Although two test functions $f_\pm$ yield the same approximate solution, they have different true solutions.  In particular,
\begin{align*}
\varepsilon &\ge \max \bigl\{\DHJRnorm[\calG]{S(f_+) - A^*(f_+,\varepsilon)}, \DHJRnorm[\calG]{S(f_-) - A^*(f_-,\varepsilon)} \bigr\} \\
&\ge \frac 12 \bigl[\DHJRnorm[\calG]{S(f_+) - A^*(f,\varepsilon)} + \DHJRnorm[\calG]{S(f_-) - A^*(f,\varepsilon)}  \bigr] \\
&\qquad \qquad \text{since } A^*(f_\pm,\varepsilon) = A^*(f,\varepsilon) \\
&\ge \frac 12 \DHJRnorm[\calG]{S(f_+) - S(f_-)} \quad \text{by the triangle inequality}\\
&\ge \frac 12 \DHJRnorm[\calG]{S(f_+ - f_-)} \quad \text{since $S$ is linear}\\
&= \eta \DHJRnorm[\calG]{S(u)}.
\end{align*}
Thus, we have \yuhannote{Please check the proof below}
\begin{align*}
\varepsilon^2  & \ge \eta^2 \DHJRnorm[\calG]{S(u)}^2= 
\eta^2 \sum_{k=0}^{j} \DHJRbignorm[\calG]{S(u^{(k)})}^2  \frac{b^{2(k-j)}}{\lambda_{n_k}^2}\\
& \ge \eta^2 
\sum_{k=0}^{j} \DHJRbignorm[\calF]{u^{(k)}}^2 
b^{2(k-j)} 
\qquad {\color{red} \text{since } \DHJRbignorm[\calG]{S(u^{(k)})} \ge \lambda_{n_k} \DHJRbignorm[\calF]{u^{(k)}} }\\
& \ge  \eta^2 b^{2(k^{*}-j)} \qquad \text{ where } k^* = \argmax_{0 \le k \le j} \DHJRbignorm[\calF]{u^{(k)}} \\
& \ge \eta^2 = \frac{(a-1)^2c^2}{(a+1)^2R^2} 
\qquad {\color{red} \text{since } b < 1}\\
&
=\frac{(a-1)^2 \rho^2}{(a+1)^2R^2}\left[\left(1+\frac{(a-1)^2}{(a+1)^{2} R^2 }\right)\sum_{k=0}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}\right]^{-1} 
\qquad {\color{red} \text{by \eqref{DHJRetaDef}}}\\
& =  \rho^2 \left[\left(\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right)\sum_{k=0}^j \frac{b^{2(k-j)}}{\lambda_{n_{k}}^2}\right]^{-1} \\
& = \rho^2 \left[\left(\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right) b^2F(j+1)\right]^{-1} 
\qquad {\color{red} \text{by \eqref{jsum}}}.
\end{align*}
{\color{red} This inequality is equivalent to 
\begin{equation}
    F(j+1) \ge  \frac{\rho^2}{\varepsilon^2 b^2}
\left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right]^{-1}.
\end{equation}

This lower bound must be satisfied by $j$ to be consistent with the assumption $\textup{comp}(\calA(\calC),\varepsilon,\rho) \le n_j - 2$.  Thus, any $j$ violating this inequality satisfies $\textup{comp}(\calA(\calC),\varepsilon,\rho) \ge n_j - 1$.  Since $F$ is strictly increasing, the largest $j$ violating this inequality must satisfy
\begin{equation}
F(j+1) < \frac{\rho^2}{\varepsilon^2 b^2}
\left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right]^{-1} \le F(j+2),
\end{equation}
which is the definition of  $j^\ddagger$ above in the statement of this theorem.  This proves the lower bound on $\textup{comp}(\calA(\calC),\varepsilon,\rho)$. }
\end{proof}

The next step is to show that the cost of our algorithm is essentially no worse than that of the optimal algorithm.
\yuhannote{The proof of theorem 3 needs to be checked.}
\begin{theorem}
\label{thm:CostNoWorse}
Under assumption \eqref{lambdaRatio} $\textup{cost}(\widetilde{A},\calC, \varepsilon,\rho)$ is essentially no worse than $\textup{comp}(\calA(\calC),\varepsilon,\rho)$. 
\end{theorem}
\begin{proof} 
{\color{red} We need an inequality relating $F(j)$, which appears in the definition of $j^\dagger$ in Theorem \ref{thm:compcost} and $F(j+3)$, which is related to the definition of $j^\ddagger$ in Theorem \ref{thm:lowbdcomp}.  According to \eqref{lambdaRatio} and the definition of $F$ in \eqref{jsum}, for any $\ell \in \mathbb{N}$,
\begin{equation*}
R^{2\ell} F(j) \ge R^{2\ell} \frac{b^{-2}}{\lambda_{n_{j-1}}^2} \geq \frac{b^{-2}}{\lambda_{n_{j+\ell-1}}^2},
\end{equation*}
and so
\begin{align*}
b^{-2\ell}[1 + b^2R^2\cdots + (b^2R^2)^\ell] F(j) & \ge F(j)b^{-2\ell} + \frac{b^{-2\ell}}{\lambda_{n_{j}}^2} + \cdots + \frac{b^{-2}}{\lambda_{n_{j+\ell-1}}^2} \\
& = \sum_{j=0}^{j-1} \frac{b^{2(k - j -\ell)}}{\lambda_{n_{k}}^2} + \frac{b^{-2\ell}}{\lambda_{n_{j}}^2} + \cdots + \frac{b^{-2}}{\lambda_{n_{j+\ell-1}}^2} \\
& = F(j + \ell).
\end{align*}
The $\ell=3$ case of this inequality implies the following lower bound on $F(j)$ in terms of $F(j+3)$:
\begin{equation*} 
F(j) \ge  \frac{b^6 F(j+3)}{1 + b^2R^2 + b^4R^4 + b^6R^6} \\
=  \frac{\omega^2 a^4 b^2}{1 - b^2}\left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right] F(j+3),
\end{equation*}
where
\begin{equation} \label{newomegadef}
\omega = \sqrt{\frac{(1 - b^2)b^4}{a^4(1 + b^2R^2 + b^4R^4 + b^6R^6)}\left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right]^{-1}},
\end{equation}
Note that $\omega$ does not depend on $\rho$ or $\varepsilon$ but only on the definition of $\calC$.

For any positive $\rho$ and $\varepsilon$, let $j^\dagger = j^\dagger(\varepsilon,\rho)$ be defined as in Theorem \ref{thm:compcost} and let $j^\ddagger = j^\ddagger(\varepsilon,\rho)$ be defined as in Theorem \ref{thm:lowbdcomp}.  Then by those Theorems, $j^\dagger(\varepsilon,\rho)$ can be bounded above in terms of $ j^{\ddagger}(\omega\varepsilon,\rho)$ for $\omega$ defined in \eqref{newomegadef}:
\begin{align*} 
j^\dagger(\varepsilon,\rho) &\le \min \left \{j \in \bbN : F(j) \ge \frac{\rho^2 a^4}{\varepsilon^2 (1-b^2)}   \right\} \\
& \le \min \left \{j \in \bbN : \frac{\omega^2 a^4 b^2}{1 - b^2}\left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right] F(j+3) \ge \frac{\rho^2 a^4}{\varepsilon^2 (1-b^2)}    \right\} \\
& = \min \left \{j \in \bbN :  F(j+3) \ge \frac{\rho^2}{\omega^2 \varepsilon^2 b^2} \left[\frac{(a+1)^{2} R^2 }{(a-1)^2} + 1\right]^{-1} \right\} \\
& = j^{\ddagger}(\omega\varepsilon,\rho) - 1.
\end{align*}
By the argument above, it follows that $ j^\dagger(\varepsilon,\rho) \le j^\ddagger(\omega\varepsilon,\rho) -1$, and again by Theorems \ref{thm:compcost} and  \ref{thm:lowbdcomp} it follows that 
\[
\textup{cost}(\widetilde{A},\calC,\varepsilon,\rho)
= n_{j^\dagger}(\varepsilon,\rho)
\le n_{j^{\ddagger}(\omega\varepsilon,\rho)-1} 
\le n_{j^{\ddagger}(\omega\varepsilon,\rho)} -1 
\le \textup{comp}(\calA(\calC),\omega\varepsilon,\rho) .
\]
Therefore, our algorithm is essentially no more costly than the optimal algorithm. }
\end{proof}
